Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-10-03 submission/NatCom_all_models.tex
pass to address most of the presubmission comments for nat comm
minor changes before submit to Nat Comm

We summarize results from a large-scale analysis of hundreds of publicly-available  models.
This broader analysis on a much larger set of CV and NLP models, with a more diverse set of architectures, and developed for a wider range of tasks, complements the previous more detailed analysis on CV and NLP models.
See Appendix~\ref{sxn:appendix} for details.
To quantify the relationship between quality metrics and the reported test error and/or accuracy metrics, we use ordinary least squares to regress the metrics on the Top1 (and Top5) reported errors (as dependent variables).
These include Top5 errors for the ImageNet-1K model, percent error for the CIFAR-10/100, SVHN, CUB-200-2011 models, and Pixel accuracy (Pix.Acc.) and Intersection-Over-Union (IOU) for other models.
We regress them individually on each of the norm-based and PL-based metrics.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|}
%KDD% \begin{tabular}{|p{0.75in}|c|c|c|c|}
\hline
%    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
%    & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\langle\hat{\alpha}=\alpha\log\lambda_{max}\rangle$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
Series        & $\log\Vert\cdot\Vert_{F}$ & $\log\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\log\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
\hline
$R^{2}$ (mean) & 0.63 &  0.55 & \textbf{0.64} & \textbf{0.64} \\
$R^{2}$ (std)  & 0.34 &  0.36 & \textbf{0.29} &          0.30 \\
\hline
$MSE$ (mean)   & 4.54 &  9.62 &          3.14 & \textbf{2.92} \\
$MSE$ (std)    & 8.69 & 23.06 &          5.14 & \textbf{5.00} \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of linear regression fits for different average Log Norm and Weighted Alpha metrics across 5 CV datasets, 17 architectures, covering 108 (out of over 400) different pretrained DNNs.  
         We include regressions only for architectures with five or more data points, and which are positively correlated with test error.
         These results can be readily reproduced using the Google Colab notebooks. (See Appendix~\ref{sxn:appendix} for details.) 
        }
\label{table:results}
\end{table}

>>>>

We summarize results from a large-scale analysis of hundreds of publicly-available  models.
This broader analysis on a much larger set of CV and NLP models, with a more diverse set of architectures, and developed for a wider range of tasks, complements the previous more detailed analysis on CV and NLP models.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
To quantify the relationship between quality metrics and the reported test error and/or accuracy metrics, we use ordinary least squares (similar results are seen with rank correlation metrics) to regress the metrics on the Top1 (and Top5) reported errors (as dependent variables).
These include Top5 errors for the ImageNet-1K model, percent error for the CIFAR-10/100, SVHN, CUB-200-2011 models, and Pixel accuracy (Pix.Acc.) and Intersection-Over-Union (IOU) for other models.
We regress them individually on each of the norm-based and PL-based metrics.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|}
\hline
%    & Frobenius Norm & Spectral Norm & Weighted Alpha & Alpha-Norm \\
%    & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\langle\hat{\alpha}=\alpha\log\lambda_{max}\rangle$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
Series        & $\log\Vert\cdot\Vert_{F}$ & $\log\Vert\cdot\Vert_{\infty}$ & $\hat{\alpha}$ & $\log\Vert\cdot\Vert^{\alpha}_{\alpha}$ \\
\hline
$R^{2}$ (mean) & 0.63 &  0.55 & \textbf{0.64} & \textbf{0.64} \\
$R^{2}$ (std)  & 0.34 &  0.36 & \textbf{0.29} &          0.30 \\
\hline
$MSE$ (mean)   & 4.54 &  9.62 &          3.14 & \textbf{2.92} \\
$MSE$ (std)    & 8.69 & 23.06 &          5.14 & \textbf{5.00} \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of linear regression fits for different average Log Norm and Weighted Alpha metrics across 5 CV datasets, 17 architectures, covering 108 (out of over 400) different pretrained DNNs.  
         We include regressions only for architectures with five or more data points, and which are positively correlated with test error.
         These results can be readily reproduced using the Google Colab notebooks. 
         (See the Supplementary Information
         %~\ref{sxn:appendix} 
         for details.) 
        }
\label{table:results}
\end{table}

### 2020-10-03 submission/NatCom_overall_approach.tex
minor changes before submit to Nat Comm

Consider the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, and weight matrices $\mathbf{W}_{l}$ and bias vectors $\mathbf{b}_{l}$, as 
the minimization of a general Loss Function $\mathcal{L}$ over the training data instances and labels, $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$.
For a typical supervised classification problem, the goal of training is to construct $\mathbf{W}_{l}$ and $\mathbf{b}_{l}$ that capture correlations in the data, in the sense of solving
\begin{equation}
\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N}\left(\mathcal{L}(E_{DNN}(\mathbf{x}_{i})-y_{i})\right)  ,
\end{equation}
where the loss function $\mathcal{L}(\cdot)$ can take on a myriad of forms~\cite{JC17_TR}, and where the Energy (or optimization) Landscape function
\begin{equation}
%E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})
E_{DNN} = f(\mathbf{W}_{1},\ldots,\mathbf{W}_{L},\mathbf{b}_{1},\ldots,\mathbf{b}_{L})
\label{eqn:dnn_energy}
\end{equation}
depends on the weights and biases.
\michael{See comments for math equations.}
For a trained model, $E_{DNN}$ does not explicitly depend on the data, but rather it maps data instance vectors ($\mathbf{x}_i$ values) to predictions ($y_{i}$ labels).
\michael{Mark found that sentence confusing.}
Therefore, one can analyze $E_{DNN}$ in the absence of any training or test~data.

@@

Our approach involves analyzing individual DNN weight matrices, for (depending on the architecture) fully-connected and/or convolutional layers.
Each DNN layer contains one or more layer 2D  $N_{l}\times M_{l}$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, e.g., extracted from 2D Convolutional layers, where $N > M$.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
(We may drop the $i$ and/or $i,l$ subscripts below.)
The best performing quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's Empirical Correlation Matrix, $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
We consider the following metrics:
\begin{eqnarray}
& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
& & \text{$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
\end{eqnarray}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
These eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, $\lambda_{i}=\sigma^{2}_{i}$.
All four metrics can be computed easily from DNN weight matrices.
The first two metrics are well-known in ML.
The last two metrics deserve special mention, as they depend on an empirical parameter $\alpha$ that is the PL exponent that arises in the recently-developed Heavy Tailed Self Regularization (HT-SR) Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.

>>>>

Consider the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, and weight matrices $\mathbf{W}_{l}$ and bias vectors $\mathbf{b}_{l}$, as 
the minimization of a general loss function $\mathcal{L}$ over the training data instances and labels, $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$.
For a typical supervised classification problem, the goal of training is to construct (or learn) $\mathbf{W}_{l}$ and $\mathbf{b}_{l}$ that capture correlations in the data, in the sense of solving
\begin{equation}
\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N}\left(\mathcal{L}(E_{DNN}(\mathbf{x}_{i})-y_{i})\right)  ,
\end{equation}
where the loss function $\mathcal{L}(\cdot)$ can take on a myriad of forms~\cite{JC17_TR}, and where the energy (or optimization) landscape function
\begin{equation}
%E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})
E_{DNN} = f(\mathbf{x}_{i} ; \mathbf{W}_{1},\ldots,\mathbf{W}_{L},\mathbf{b}_{1},\ldots,\mathbf{b}_{L})
\label{eqn:dnn_energy}
\end{equation}
depends parametrically on the weights and biases.
\michael{See comments for math equations.}
For a trained model, $E_{DNN}$ does not explicitly depend on the data, but rather it maps data instance vectors ($\mathbf{x}_i$ values) to predictions ($y_{i}$ labels).
\michael{Mark found that sentence confusing.}
Therefore, one can analyze $E_{DNN}$ in the absence of any training or test~data.

@@

Our approach involves analyzing individual DNN weight matrices, for (depending on the architecture) fully-connected and/or convolutional layers.
Each DNN layer contains one or more layer 2D  $N_{l}\times M_{l}$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, e.g., extracted from 2D Convolutional layers, where $N > M$.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
(We may drop the $i$ and/or $i,l$ subscripts below.)
The best performing quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's empirical correlation matrix, $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
We consider the following metrics:
\begin{eqnarray}
& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
& & \text{$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
\end{eqnarray}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
These eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, $\lambda_{i}=\sigma^{2}_{i}$.
All four metrics can be computed easily from DNN weight matrices.
The first two metrics are well-known in ML.
The last two metrics deserve special mention, as they depend on an empirical parameter $\alpha$ that is the PL exponent that arises in the recently-developed Heavy Tailed Self Regularization (HT-SR) Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.
